---
title: "mL09_BasicDataTypes"
author: "David M Hannon"
date: "27/12/2021"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: true
---

# Vector Creation

Create double variables
```{r}
double_var <- c(2, 3.4, 7, 0, 9.99)
```

Integer variables

```{r}
integer_var <- c(1L, 5L, 82L)
```

Logical variables

```{r}
logic_var <- c(TRUE, FALSE, T, F)
```

Character variable

```{r}
character_var <- c('a', 'b', 'hello', '86')
```

# Types of vector (vector checking)

Use the `typeof` command to check what type of data a vector is.

```{r}
typeof(character_var)
typeof(logic_var)
```

Use `is.<VECTOR>` to check using a **logical output**.

```{r}
is.numeric(character_var)
is.integer(integer_var)
is.numeric(integer_var)   # integer and dbl are both numeric
```

# Type coercion

If a vector has numerous data types within it, R will coerce them all to the 
most flexible type that fits.

**NB: logical > integer > numeric > complex > character**

```{r}
typeof(c(1L, 2L, 4))   # it is a double because they can ALL fit this data type 
```

You can cast one as another.

```{r}
as.integer(double_var)   # observe how it disregards all after a decimal point

# the above can also be nested in another function
as.logical(as.integer(double_var))
```

R will convert **if it can**. Otherwise, an NA is generated.

```{r}
as.numeric(character_var)
```

