---
title: "Data Frames"
author: "David Hannon"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

# Introduction

This tutorial will help you to understand

- how to create a dataframe

- how to access elements in a dataframe

- how to delete columns from a dataframe

## Dataframe Creation

A dataframe is created either by importing data, or from hand.

Nowadays tibbles replace dataframes. They are dataframes, but they print nicer, 
and have other improvements.

```{r}
library(tidyverse)
```

```{r}
wind_oem <- tibble(company = c("Vestas", "SGRE", "GE"),
                       market.share = c(16, 15, 10))
wind_oem
```

## Access Elements

You can access columns with $ operator

```{r}
wind_oem$name
```

or with the second argument of [] operator

```{r}
wind_oem[, 1]
```

or by specifying the name of the column (this is the same as the Python way).

```{r}
wind_oem['company']
```


The same way you can access rows.

```{r}
wind_oem[1, ]
```


## Delete Columns

If you want to delete columns you can assign NULL operator to the column.

```{r}
wind_oem$market_share <- NULL
```

