---
title: "Univariate Linear Regression"
author: "David Hannon"
date: "January 20th 2022"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Understanding

We will work with a dataset that is provided with the **dplyr** package. Based on Star Wars characters, we will predict a person's weight based on their height.

What is an independent and dependent variable? Well, the height is largely defined in our genes, but our weight can be more easily influenced. So, height will be the independent variable and mass the dependent variable.

## Packages

We load required packages.

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
```

## Data Import

We load the data.

```{r}
data("starwars")
set.seed(123)
```

## Visualisation

We write a small function for visualising the data.

```{r}
plot_height_mass <- function(df) {
  g <- ggplot(df, aes(x = height, y = mass))
  g <- g + geom_point()
  g <- g + geom_smooth(method="lm", se=F)
  g
}

plot_height_mass(starwars)
```

# Data Preparation

There is a clear outlier we need to handle.

```{r}
starwars %>% 
  filter(mass > 1000) %>% 
  select(name, height, mass)
```


```{r}
starwars_filt <- starwars %>% 
  filter(mass < 1000)
         
plot_height_mass(starwars_filt)
```

# Modeling

Model is created with *lm()*.

```{r}
model <- lm(data = starwars_filt, 
            formula = mass ~ height)
```

We take a look at the summary.

```{r}
model_summary <- model %>% 
  summary()

model_summary
```

R**2 is calculated.

```{r}
model_summary$r.squared
```

# Predictions

```{r}
starwars_filt$mass_predictions <- predict(object = model, 
                                         newdata = starwars_filt)
```

Now we can visualise the result.

```{r}
g <- plot_height_mass(starwars_filt)
g <- g + geom_point(data = starwars_filt, 
                    aes(x = height, 
                        y = mass_predictions),
                    col = "red", alpha = .3)
g
```

We see that our predictions exactly match the smoothed line from ggplot.

But how good do they match our actual data.

```{r}
g <- ggplot(data = starwars_filt,
            aes(x = mass,
                y = mass_predictions,
                label = name)
            )
g <- g + geom_point()
g <- g + geom_text(size = 3)
g <- g + geom_abline(slope = 1, 
                     intercept = 0,
                     col = "blue")
g <- g + labs(x = "Actual Mass [lb]",
              y = "Predicted Mass [lb]",
              title = "Mass Correlation")
g
```
